---
title: "Case study 1"
author: "stephanie"
date: "2022-11-04"
output: 
  html_document: 
    toc: yes
---

# DIVVY CASE STUDY DATA ANALYSIS USING R STUDIO

##This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman (found here: https://artscience.blog/home/divvy-dataviz-case-study).

##The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: “In what ways do members and casual riders use Divvy bikes differently?”

### Loading R packages
notes: setting r environment by installing and loading `tidyverse`for data import and wrangling,`ggplot2`for data visualization and `lubridate`or date functions


```{r loading packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```


### To display and set working directory
notes:`getwd()` #displays your working directory
`setwd("/Users/kevinhartman/Desktop/Divvy_Exercise/csv")`#sets your working directory to simplify calls to data ... make sure to use your OWN username instead of mine ;)

```{r display working directory, had already set working directory}
getwd()

```


## STEP 1 COLLECT DATA
notes:upload divvy data set using `read_csv`

```{r upload data set, message=FALSE, warning=FALSE}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```


## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
notes: combine data set using 
```{r see column names}
colnames(q4_2019)
colnames(q3_2019)
colnames(q2_2019)
colnames(q1_2020)
```


###Rename columns  to make them consistent with q1_2020 (as this will be the supposed going-forward table design for Divvy)

```{r rename columns to match q1_2020, message=FALSE, warning=FALSE}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))
```


###Convert ride_id and rideable_type to character so that they can stack correctly

```{r convert to character}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```


###Stack individual quarter's data frames into one big data frame

```{r combine rows}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```


###Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020

```{r drop columns}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```


## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
notes: list col names`colnames()`,count rows`nrow()`,check data dimensions`dim()`,see first 6 rows of data frame`head()``tail()`,data types`str()`,statistic summary`summary`

```{r Clean dataset, message=FALSE, warning=FALSE}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```


##PROBLEMS TO FIX
###(1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.
#### Reassign to the desired values (we will go with the current 2020 labels then check again to see if the observations were reasigned

```{r see observations in this column, message=FALSE, warning=FALSE, include=FALSE, paged.print=TRUE}
all_trips$member_casual
all_trips<-all_trips %>% 
  mutate(member_casual=recode(member_casual,
         "Subscriber"="member",
         "Customer"="casual"))
all_trips$member_casual
```

####(2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.
###https://www.statmethods.net/input/dates.html more on date formats in R found at that link.The default format for date in r is yyyy-mm-dd

```{r format date}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

###(3) We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
###https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html

```{r add calculated field}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

```

###Inspect the structure of the columns
###Convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r change data type}
str(all_trips)
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```


###(4) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides
###We will create a new version of the dataframe (v2) since data is being removed
###https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/

```{r remove negative data}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
## STEP 4 CONDUCT DATA ANALYSIS
###DESCRIPTIVE STATISTICS(# You can condense the four lines above to one line using summary() on the specific attribute`summary(all_trips_v2$ride_length)`
)

```{r descriptive statistics}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride

```

###Compare members and casual users

```{r summary by member_casual}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
###See the average ride time by each day for members vs casual users
```{r aggregate ride time by member_casual}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

###Notice that the days of the week are out of order. Let's fix that.
```{r fix days of the week}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

###Now, let's run the average ride time by each day for members vs casual users
```{r repeat agrregation of ride time by member_casual}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

###analyze ridership data by type and weekday

```{r summary,message=FALSE, warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%                  #groups by usertype and weekday
  summarise(number_of_rides = n()						          	#calculates the number of rides and                                                            average duration 
  ,average_duration = mean(ride_length)) %>% 	        	# calculates the average duration
  arrange(member_casual, weekday)							         	# sorts

```


###Let's visualize the number of rides by rider type

```{r visualize by number of rides, message=FALSE, warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at,label=TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x=weekday, y= number_of_rides,fill=member_casual))+
  geom_col(position="dodge")
            
```


### Let's create a visualization for average duration

```{r visualize by avarage time, message=FALSE, warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```


## STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
###Create a csv file that we will visualize in Excel, Tableau, or my presentation software
###You can read more here: https://datatofish.com/export-dataframe-to-csv-in-r/

```{r export summary file}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'C:/Users/shari/OneDrive/Documents/Projectone/avg_ride_length.csv')
```












